Demystifying Votes: A Hierarchical Bayesian Model for Predicting the 2024 US Presidential Election

### 2024-11-02 paper/paper.qmd
for prose, all words except "insight" and "reveal" removed or replaced by proper words

Once the survey is designed, polling begins, and responses are aggregated by the polling organization. Respondents' answers are recorded and categorized, converting their opinions into structured data points, though these lack the nuances of human complexity. To protect privacy, data are often anonymized and aggregated, retaining key insights. After collection, raw data undergo cleaning and processing to minimize sampling biases and errors. Yet, any single poll can have random and systematic error sources, so combining multiple poll results helps mitigate these errors [@blumenthal2010polls].

@@

**state/national**: Political environments and voter preferences may vary significantly by state, and national polls cannot fully reflect conditions in individual states. The electoral college system in the U.S. makes state-level approval rates crucial for predicting the likelihood of winning.

@@

Although low-frequency observations may introduce some bias, directly removing them would result in information loss, especially in election predictions where the representativeness of different states and pollsters is crucial. Different pollsters and states have their own resources and priorities, leading to uneven data coverage. Some pollsters may prefer to publish data at specific times, while certain states, due to smaller populations or stable voter behavior, may receive less attention. Thus, it is challenging to completely balance the data quantity across different items. Although some observations have low frequency, they still contain valuable information. For instance, even though smaller or remote states may have limited data, their voter preferences are still significant under the Electoral College system. Completely deleting these observations would diminish these states' representation in the overall analysis, weakening the model's accuracy in reflecting support rates. Additionally, we aim to cover as many pollsters and states as possible with the model, so we retain even low-frequency observations. Thus, we keep these observations and mitigate the bias they introduce by combining other variables to build a more comprehensive and robust predictive model.

@@

We implemented the model using the rstanarm package, leveraging MCMC sampling for Bayesian inference. To validate the model, we performed out-of-sample testing and calculated the RMSE (Root Mean Square Error), which resulted in 2.31. In the context of complex, socially influenced political surveys, this error level is considered acceptable. Convergence diagnostics in @fig-post4 in [Appendix -@sec-model-details] show that R-hat values are close to 1, and the Effective Sample Size (ESS) is generally large (often > 400), indicating good independence among samples. Sensitivity analyses on priors indicate that slight adjustments do not significantly alter the predictions, demonstrating model stability.

@@

At the state level, @fig-results shows predictions indicate significant variations between states. For example, support in Texas is considerably higher than in California, with a gap of over 10 percentage points, while swing states like Pennsylvania and Michigan display greater volatility. This result further validates the model’s robustness; the introduction of a random intercept for each state reflects systematic differences across states. This design aligns with data trends, suggesting that states with diverse geographic and demographic profiles exhibit unique political tendencies.

@@

To draw final conclusions from these predictions, it is essential to consider historical polling data and past election results. In previous U.S. presidential elections, shifts in sentiment in swing states have played a crucial role in determining the outcome. For example, the 2016 election highlighted the importance of swing states like Pennsylvania, Michigan, and Wisconsin, where polls underestimated Trump’s support. A study by the Pew Research Center noted that last-minute sentiment shifts and polling errors led to an unexpected advantage for Trump [@pew2021pollinglimitations]. Studies by *The Economist* emphasize the critical impact of swing states on election outcomes, also noting that national polls do not always reflect the dynamics of the Electoral College [see @economist2020electionforecast]. The Electoral College system can result in a candidate winning the popular vote but losing the election, or vice versa. This system has affected U.S. elections multiple times, demonstrating the importance of state-level results relative to national polling trends.

@@

#### The Randomness of the Real World
Voter preferences can shift rapidly due to events like debates, economic changes, or international incidents, challenging election predictions. Regular polling may fail to capture these shifts if not updated frequently, leading to outdated and potentially misleading forecasts. The [Harvard Data Science Review](https://hdsr.mitpress.mit.edu/pub/nw1dzd02/release/2) highlighted this uncertainty in the 2020 election, where state-level opinion volatility was influenced by economic and educational factors, necessitating models that incorporate these dynamic interactions [@gelman2020regression].

>>>>

Once the survey is designed, polling begins, and responses are aggregated by the polling organization. Respondents' answers are recorded and categorized, converting their opinions into structured data points, though these lack of human complexity. To protect privacy, data are often anonymized and aggregated, retaining key insights. After collection, raw data undergo cleaning and processing to minimize sampling biases and errors. Yet, any single poll can have random and systematic error sources, so combining multiple poll results helps mitigate these errors [@blumenthal2010polls].

@@

**state/national**: Political environments and voter preferences may vary significantly by state, and national polls cannot fully reflect conditions in individual states. The electoral college system in the U.S. makes state-level approval rates important for predicting the likelihood of winning.

@@

Although low-frequency observations may introduce some bias, directly removing them would result in information loss, especially in election predictions where the representativeness of different states and pollsters is important. Different pollsters and states have their own resources and priorities, leading to uneven data coverage. Some pollsters may prefer to publish data at specific times, while certain states, due to smaller populations or stable voter behavior, may receive less attention. Thus, it is challenging to completely balance the data quantity across different items. Although some observations have low frequency, they still contain valuable information. For instance, even though smaller or remote states may have limited data, their voter preferences are still significant under the Electoral College system. Completely deleting these observations would diminish these states' representation in the overall analysis, weakening the model's accuracy in reflecting support rates. Additionally, we aim to cover as many pollsters and states as possible with the model, so we retain even low-frequency observations. Thus, we keep these observations and mitigate the bias they introduce by combining other variables to build a more robust predictive model.

@@

We implemented the model using the rstanarm package, using MCMC sampling for Bayesian inference. To validate the model, we performed out-of-sample testing and calculated the RMSE (Root Mean Square Error), which resulted in 2.31. In the context of complex, socially influenced political surveys, this error level is considered acceptable. Convergence diagnostics in @fig-post4 in [Appendix -@sec-model-details] show that R-hat values are close to 1, and the Effective Sample Size (ESS) is generally large (often > 400), indicating good independence among samples. Sensitivity analyses on priors indicate that slight adjustments do not significantly alter the predictions, demonstrating model stability.

@@

At the state level, @fig-results shows predictions indicate significant variations between states. For example, support in Texas is considerably higher than in California, with a difference of over 10 percentage points, while swing states like Pennsylvania and Michigan display greater volatility. This result further validates the model’s robustness; the introduction of a random intercept for each state reflects systematic differences across states. This design aligns with data trends, suggesting that states with diverse geographic and demographic profiles exhibit unique political tendencies.

@@

To draw final conclusions from these predictions, it is essential to consider historical polling data and past election results. In previous U.S. presidential elections, shifts in sentiment in swing states have played a important role in determining the outcome. For example, the 2016 election highlighted the importance of swing states like Pennsylvania, Michigan, and Wisconsin, where polls underestimated Trump’s support. A study by the Pew Research Center noted that last-minute sentiment shifts and polling errors led to an unexpected advantage for Trump [@pew2021pollinglimitations]. Studies by *The Economist* emphasize the critical impact of swing states on election outcomes, also noting that national polls do not always reflect the dynamics of the Electoral College [see @economist2020electionforecast]. The Electoral College system can result in a candidate winning the popular vote but losing the election, or vice versa. This system has affected U.S. elections multiple times, demonstrating the importance of state-level results relative to national polling trends.

@@

#### The Randomness of the Real World
Voter preferences can shift due to events like debates, economic changes, or international incidents, challenging election predictions. Regular polling may fail to capture these shifts if not updated frequently, leading to outdated and potentially misleading forecasts. The [Harvard Data Science Review](https://hdsr.mitpress.mit.edu/pub/nw1dzd02/release/2) highlighted this uncertainty in the 2020 election, where state-level opinion volatility was influenced by economic and educational factors, necessitating models that incorporate these dynamic interactions [@gelman2020regression].
